\vspace{-2cm} Inference on the Effects of Observed Features \\ in Latent Space Models for Networks

### 2018-02-16 manuscript_rearranged.tex
making revisions to the intro

\noindent Due to the complex interdependence found in network data,
networks scholars draw upon an increasingly sophisticated
toolkit for statistical inference. The Latent space model (LSM) for
network data combines a conventional regression model with an embedding of actors in a latent space. In the LSM, the expected relationship between two actors is dependent on observed covariates and a function of their positions in the latent space. In numerous applications, researchers have assumed
that the latent spatial embedding can control for latent
confounding structure. Latent confounding structure includes unmeasured actor attributes that can be represented spatially, as well as network dependencies such as reciprocity and transitivity. There has been little research that considers the LSM's
performance in adjusting for latent confounding structure. We investigate the
LSM's performance via a simulation study. In the presence of an
unmeasured actor attribute that can be modeled perfectly using a latent space,
we find that the LSM exhibits unacceptable levels of bias and Type 1 error for inferences on measured covariates. However, the prediction error of the LSM when latent confounding structure is present is substantially lower in most cases. We
conclude that the LSM is most appropriate for exploration or
prediction, unless it can be assumed that all potential confounders have been measured.
%\footnote{This work was supported by National Science Foundation
%grants DGE-1144860, SES-1558661, SES-1619644, and CISE-1320219. Any
%opinions, findings, and conclusions or recommendations are those of
%the authors and do not necessarily reflect those of the sponsor.}
\end{abstract}
\thispagestyle{empty}
% \doublespacing
% Description of the possible challenges
\section{Introduction}

Inferential analysis of political network data has grown increasingly
sophisticated in recent years. Political networks scholars are well
versed in the risks associated with ignoring unmodeled network
structure. Dependencies such as reciprocity, transitivity, and
homophily---if not accounted for---can lead to biased estimates and
errors in hypothesis testing, much in the way that omitted variable
bias can affect results in conventional regression models
\citep{ward2007disputes,kinne2014,cranmerisq,hays2010spatial}. A
number of statistical modeling frameworks have been proposed to
account for confounding structure in network data that cannot be
modeled with observed covariates. These include the exponential random
graph model (ERGM) \citep[e.g.,
][]{lazer2010,cranmer2011pa,desmarais2012psj}, the latent space model
(LSM) \citep[e.g.,
][]{ward2007disputes,ward2007persistent,kirkland2012multimember}, and
the stochastic actor oriented model (SAOM) \citep[e.g.,
][]{berardo2010ajps,kinne2014}.

The approach to adjusting for dependencies in the ERGM and SAOM is
quite similar to adjusting for confounding covariates in regression
modeling. The researcher specifies a set of dependencies that (s)he
hypothesizes to be important in the generative model for the
network. These dependencies are then explicitly included in a model
that simultaneously represents the effects of observed covariates
\citep{cranmer2011pa}. The LSM takes a different approach, which
involves the incorporation of latent variables to model network
structure. The LSM has an advantage over ERGM and SAOM in that
researchers need not develop a set of hypothesized dependencies in
order to model network structure that is not reflected in observed
covariates. However, this advantage hinges upon the capacity for the
LSM to discover unmodeled structure that could otherwise be attributed
to the observed covariates.

Despite their growing popularity, few studies exist that investigate
the performance of these models in adjusting for confounding network
structure. In the current study, we focus on the LSM, examining its
performance in reducing estimation and inferential errors regarding
the effects of observed covariates via adjustment for confounding
network structure. There are two reasons that using the LSM may lead
to increased error---one may result in Type 1 inferential error, and
the other in Type 2 error. First, if the unobserved (i.e., latent)
network structure is truly correlated with the observed covariates,
the unobserved structure that can be correlated with the observed
variable may be attributed to the observed variable. Under this
condition the latent space parameters are used to model sources of
variation that cannot be attributed to the observed covariate, and
inferences regarding the observed covariates remain subject to omitted
variable bias, which leads to an inflated Type 1 inferential error
rate. Second, the latent configurations inferred may result in a
representation of the network wherein a node's position in the latent
space is spuriously correlated with the observed covariates. Inferring
latent variables that are correlated with observed covariates would
lead to reduced efficiency in estimating observed covariate effects,
and result in a high Type 2 error rate.

>>>>

\noindent Due to the complex interdependence found in network data,
networks scholars draw upon an increasingly sophisticated
toolkit for building models of networks. The Latent space model (LSM) for
network data combines a conventional regression model with an embedding of actors in a latent space. In the LSM, the expected relationship between two actors is dependent on observed covariates and a function of their positions in the latent space. In numerous applications, researchers have assumed
that the latent spatial embedding can control for latent
confounding structure. latent network structure includes unmeasured actor attributes that can be represented spatially, as well as network dependencies such as reciprocity and transitivity. There has been little research that considers the LSM's
performance in adjusting for latent network structure when making inferences about covariates. We investigate the
LSM's performance via a simulation study. In the presence of an
unmeasured actor attribute that can be modeled perfectly using a latent space,
we find that the LSM exhibits unacceptable levels of bias and Type 1 error for inferences on measured covariates. However, the prediction error of the LSM when latent network structure is present is substantially lower in most cases. We
conclude that the LSM is most appropriate for exploratory or predictive modeling, unless it can be assumed that all potential confounders have been measured.
%\footnote{This work was supported by National Science Foundation
%grants DGE-1144860, SES-1558661, SES-1619644, and CISE-1320219. Any
%opinions, findings, and conclusions or recommendations are those of
%the authors and do not necessarily reflect those of the sponsor.}
\end{abstract}
\thispagestyle{empty}
% \doublespacing
% Description of the possible challenges
\section{Introduction}

Statistical models in which the outcome data (i.e., the dependent variable) consists of network data have proliferated, and grown increasingly
sophisticated, over the past decade. The defining feature of network data that renders conventional statistical models---namely, regression models---inappropriate, is that ties in a network exhibit complex forms of interdependence such as reciprocity, transitivity, and
homophily. If not accounted for, these complex dependencies can lead to biased estimates and
errors in hypothesis testing, much in the way that omitted variable
bias can affect results in conventional regression models
\citep{ward2007disputes,kinne2014,cranmerisq,hays2010spatial}.  We refer to network dependencies that cannot be modeled with observed covariates as {\em latent network structure}. A
number of statistical modeling frameworks have been proposed to
account for latent network structure in network data. These include, but are not limited to, the exponential random
graph model (ERGM) {\bf [UPDATE STRING CITES TO LOOK OUTSIDE POLISCI]} \citep[e.g.,
][]{lazer2010,cranmer2011pa,desmarais2012psj}, the latent space model
(LSM) \citep[e.g.,
][]{ward2007disputes,ward2007persistent,kirkland2012multimember}, and
the stochastic actor oriented model (SAOM) \citep[e.g.,
][]{berardo2010ajps,kinne2014}.

The approach to adjusting for latent network structure in the ERGM and SAOM is
quite similar to adjusting for confounding covariates in regression
modeling. The researcher specifies a set of dependencies that (s)he
hypothesizes to be important in the generative model for the
network. These dependencies are then explicitly included in a model
that simultaneously represents the effects of observed covariates
\citep{cranmer2011pa}. The LSM takes a different approach, which
involves the incorporation of latent variables to model latent network structure. The LSM has an advantage over ERGM and SAOM in that
researchers need not develop a precise set of hypothesized dependencies in
order to model latent network structure. However, this advantage hinges upon the capacity for the
LSM to differentiate the effects of latent network structure 
from those of the observed covariates.

Despite their growing popularity, few studies exist that investigate
the performance of these models in adjusting for confounding network
structure. In the current study, we focus on the LSM, examining its
performance in reducing estimation and inferential errors regarding
the effects of observed covariates via adjustment for confounding
network structure. There are two reasons that using the LSM may lead
to increased error---one may result in Type 1 inferential error, and
the other in Type 2 error. First, if the latent
network structure is truly correlated with the observed covariates,
the unobserved structure that can be correlated with the observed
variable may be attributed to the observed variable. Under this
condition the latent space parameters are used to model sources of
variation that cannot be attributed to the observed covariate, and
inferences regarding the observed covariates remain subject to omitted
variable bias, which leads to an inflated Type 1 inferential error
rate. Second, the latent configurations inferred may result in a
representation of the network wherein a node's position in the latent
space is spuriously correlated with the observed covariates (i.e., the latent space crowds out the effects of observed covariates). Inferring
latent variables that are correlated with observed covariates would
lead to reduced efficiency in estimating observed covariate effects,
and result in a high Type 2 error rate.

%HERE, ADD OVERVIEW OF OUR PAPER
